### 2020-04-05 05-translation.tex

In the following use the following conventions for denoting different categories of names:
\begin{itemize}
\item
$\mathit{decs} \defeq \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ is the set of names for decorated entities (without subscripts $_i$ and $_o$);
\item
$\mathit{ents} \defeq \{ d_i, d_o~|~ d\in \mathit{decs}\}$ is the set names for entities;
\item
$\mathit{reacts}\defeq \{r_1,...,r_{|A|}\}$ is the set of reaction names $r_j$ associated with each reaction $aj$;
\item 
$\mathit{prods}\defeq \{p_1,...,p_{|A|}\}$ is the set of names for production $p_j$, associated with  reaction $aj$.
\end{itemize}

@@

   In contrast, the processes encoding the entities, $P_{s1}$,  $\overline{P}_{s2}$, provides the

@@

Viceversa, it is immediate to define the function $\mathit{unflat}$ to rebuild the complete label from the compact string (here we exploit again the half link and block notation):
$$
\mathit{unflat}(x) \defeq\left\{
\begin{array}{ll}
^{x}_{x} & \mbox{if $x\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
^{x_i}_{x_i}\backslash^{x_o}_{x_o} & \mbox{if $x\in \mathit{decs}$}\\
\end{array}
\right.
$$
$$
\mathit{unflat}(x_1\dots x_n) \defeq\ ^{\silent}\backslash \mathit{unflat}(x_1) \backslash \dots \backslash \mathit{unflat}(x_n) \backslash_{\silent}
$$

@@

With the next proposition, we analyse the structure of a c\CNA \ 
process encoding of  a reactive process after one step transition.
In the following four statements, for brevity, we let ${\cal A} = (S, A)$ be a RS, and let  $\pi=(\gamma,\delta)$ be an extended interactive process in $A$, with $\gamma=\{C_i\}_{i\in\mathbb{N}}$ and $\delta=\{D_i\}_{i\in\mathbb{N}}$.

>>>>

In the following we use the following conventions for denoting different categories of names:
\begin{itemize}
\item
$\mathit{decs} \defeq \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ is the set of names for decorated entities (without subscripts $_i$ and $_o$);
\item
$\mathit{ents} \defeq \{ d_i, d_o~|~ d\in \mathit{decs}\}$ is the set names for entities;
\item
$\mathit{reacts}\defeq \{r_1,...,r_{|A|}\}$ is the set of reaction names $r_j$ associated with each reaction $aj$;
\item 
$\mathit{prods}\defeq \{p_1,...,p_{|A|}\}$ is the set of names for production $p_j$, associated with  reaction $aj$.
\end{itemize}

@@

   In contrast, the processes encoding the entities, $P_{s1}$,  $\overline{P}_{s2}$, provide the

@@

It is then immediate to define the function $\mathit{unflat}$ to rebuild the complete label from the compact string (here we exploit again the half link and block notation):
$$
\mathit{unflat}(x) \defeq\left\{
\begin{array}{ll}
^{x}_{x} & \mbox{if $x\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
^{x_i}_{x_i}\backslash^{x_o}_{x_o} & \mbox{if $x\in \mathit{decs}$}\\
\end{array}
\right.
$$
$$
\mathit{unflat}(x_1\dots x_n) \defeq\ ^{\silent}\backslash \mathit{unflat}(x_1) \backslash \dots \backslash \mathit{unflat}(x_n) \backslash_{\silent}
$$

@@

With the next proposition, we analyse the structure of a c\CNA \ 
process encoding of  a reactive process after one transition step.
In the following four statements, for brevity, we let ${\cal A} = (S, A)$ be a RS, and let  $\pi=(\gamma,\delta)$ be an extended interactive process in $A$, with $\gamma=\{C_i\}_{i\in\mathbb{N}}$ and $\delta=\{D_i\}_{i\in\mathbb{N}}$.

### 2020-04-05 08-conclusion.tex
small corrections
prova cambio
cambio

In this paper we have introduced a variant of the {\tt link}-calculus
where prefixes are link chains and no more single links, as it was 
briefly described in the future work section in~\cite{BBB17}.
This variant allowed us to define 
%shown 
a faithful embedding of Reaction Systems,
an emerging formalism to model computationally biochemical systems.
%into multilink processes.  
This translation shows several benefits.
{\color{red} For instance, in our paper the context works non deterministically and it is recursively defined.}
%  the context behaviour is expressed recursively and no be expressed recursively. 
Also, entity mutations can be expressed easily and different reaction systems can
communicate between them.

We have then modified the classical notion (in concurrency) of bisimulation
making it more suitable for proving properties of Reaction Systems in our
framework.
We have defined a new assertion language, which allows us to specify
the properties to be verified, and have shown that it extends Hennessy-Milner logic
to our framework.

@@

As future work, we plan to implement a prototype of our framework,
with an automatic translation from RSs to  {\tt link}-calculus. 
We believe that our work can also help to extend the framework
of RSs towards a model which can improve the communication
between different RSs. We also believe that our work can make 
possible to investigate how to apply formal techniques to prove 
properties of the
modeled systems \cite{CFHOT15,OCHF16,BBGLBH2017}.

>>>>

In this paper we have introduced a variant of the {\tt link}-calculus
where prefixes are link chains and no more single links, as it was 
briefly described in the future work section in~\cite{BBB17}.
This variant allowed us to define 
%shown 
a faithful embedding of Reaction Systems,
an emerging formalism to model computationally biochemical systems.
%into multilink processes.  
This translation shows several benefits.
For instance, in our paper the context works non deterministically and it 
is recursively defined.
%  the context behaviour is expressed recursively and no be expressed recursively. 
Also, entity mutations can be expressed easily and different reaction systems can
communicate between them.

We have then modified the classical notion (in concurrency) of bisimulation
making it more suitable for proving properties of Reaction Systems in our
framework.
We have defined a new assertion language, which allows us to specify
the properties to be verified, and have shown that it extends Hennessy-Milner logic
to our framework.
We believe that our work can make 
possible to investigate how to integrate our methodology
with other formal techniques to prove 
properties of the modeled systems \cite{CFHOT15,OCHF16,BBGLBH2017}.

@@

As future work, we plan to implement a prototype of our framework,
with an automatic translation from RSs to  {\tt link}-calculus. 
We believe that our work can also help to extend the framework
of RSs towards a model which can improve the communication
between different RSs.

### 2020-04-05 09-appendix.tex

% \begin{corollary}[completeness2]
% \label{prop:completeness}
%  Let ${\cal A} = (S, A)$ be a $rs$, and let $\pi=(\gamma,\delta)$ be a $n$-step interactive process in ${\cal A}$ with $\gamma = C_0,\dots,C_n$ and $\delta = D_0,\dots,D_n$. Let $P=[|{\cal A},\gamma|]$ its c\CNA \ translation. If we define $\pi'= (\gamma',\delta')$  with   $\gamma' = C_0',\dots,C_{n-1}'$ and $\delta' = D_0',\dots,D_{n-1}'$  as
%  $C_0'=C_w \cup D_w$, and   $C_i'=C_{i+w}$, $D_i' = D_{i+w}$ with $1 \leq i \leq n-w $, then $\exists$ $P'$ such that $P \rightarrow^w P'$ and  $P'=[|{\cal A},\gamma'|]$.  
% \end{corollary}
 \begin{proof}
 The proof proceeds by induction on the number $j$, and it is similar to the one of Corollary~\ref{corr:corr2}\\
% case $j=0$\\
% It is enough to set $C_0' = C_0 \cup D_0$ and $P' =P$.\\
% case $j=1$\\
% This case falls into the case of Proposition~\ref{prop:compl1}.\\
% case $j>1$\\
% We set $C_0'=C_w \cup D_w = C_1'' \cup D_1''$, $C_g' = C_{n-w}= C_{g+1}''$, and $D_g' = D_{n-w} = D_{g+1}''$, with $1 \leq  g < n-w$.\\
% By inductive hypothesis, it holds that $\exists$ $P''$ such that $P \xrightarrow{\link{\silent}{\silent}}^{w-1} P''$ and $P'' = [| {\cal A},\gamma'|]$.
% By Lemma~\ref{lem:compl2}.1, it exists $P'$ such that $P'' \xrightarrow{\link{\silent}{\silent}} P'$,
% and $P' = [|{\cal A},\gamma'|]$. By Lemma~\ref{lem:compl2}.2, $P'$ is unique.
%% By Lemma~\ref{lem:compl2}.2, if $\exists$ $P'$ such that $P'' \xrightarrow{\link{\silent}{\silent}} P'$, then $P'$ is 
% %unique, and by Lemma~\ref{lem:compl2}.1 $P' = [|{\cal A},\gamma'|]$. 
% And we are done. 
 \end{proof}

>>>>

% \begin{corollary}[completeness2]
% \label{prop:completeness}
%  Let ${\cal A} = (S, A)$ be a $rs$, and let $\pi=(\gamma,\delta)$ be a $n$-step interactive process in ${\cal A}$ with $\gamma = C_0,\dots,C_n$ and $\delta = D_0,\dots,D_n$. Let $P=[|{\cal A},\gamma|]$ its c\CNA \ translation. If we define $\pi'= (\gamma',\delta')$  with   $\gamma' = C_0',\dots,C_{n-1}'$ and $\delta' = D_0',\dots,D_{n-1}'$  as
%  $C_0'=C_w \cup D_w$, and   $C_i'=C_{i+w}$, $D_i' = D_{i+w}$ with $1 \leq i \leq n-w $, then $\exists$ $P'$ such that $P \rightarrow^w P'$ and  $P'=[|{\cal A},\gamma'|]$.  
% \end{corollary}
 \begin{proof}
 The proof proceeds by induction on the number $j$, and it is similar to the one of Corollary~\ref{corr:corr2}.\\
% case $j=0$\\
% It is enough to set $C_0' = C_0 \cup D_0$ and $P' =P$.\\
% case $j=1$\\
% This case falls into the case of Proposition~\ref{prop:compl1}.\\
% case $j>1$\\
% We set $C_0'=C_w \cup D_w = C_1'' \cup D_1''$, $C_g' = C_{n-w}= C_{g+1}''$, and $D_g' = D_{n-w} = D_{g+1}''$, with $1 \leq  g < n-w$.\\
% By inductive hypothesis, it holds that $\exists$ $P''$ such that $P \xrightarrow{\link{\silent}{\silent}}^{w-1} P''$ and $P'' = [| {\cal A},\gamma'|]$.
% By Lemma~\ref{lem:compl2}.1, it exists $P'$ such that $P'' \xrightarrow{\link{\silent}{\silent}} P'$,
% and $P' = [|{\cal A},\gamma'|]$. By Lemma~\ref{lem:compl2}.2, $P'$ is unique.
%% By Lemma~\ref{lem:compl2}.2, if $\exists$ $P'$ such that $P'' \xrightarrow{\link{\silent}{\silent}} P'$, then $P'$ is 
% %unique, and by Lemma~\ref{lem:compl2}.1 $P' = [|{\cal A},\gamma'|]$. 
% And we are done. 
 \end{proof}

 \theocorr*
 \begin{proof}
The proof is just an adaptation of the classical result. The two implications are proved separately.
\begin{description}
\item[$\sim_{\mathsf{F}}\ \subseteq\ \equiv_{\mathcal{L}_{\mathsf{F}}}$:]
Given any two processes $P\sim_{\mathsf{F}} Q$ we need to prove that for any bioHML formula $\mathsf{G}$ we have $P\entails \mathsf{G}$ iff $Q\entails \mathsf{G}$. 
Without loss of generality, we prove that $P\entails \mathsf{G}$ implies $Q\entails \mathsf{G}$.
The proof is by structural induction on $\mathsf{G}$.
\begin{itemize}
\item 
if $\mathsf{G}={\tt t}$, then $Q\entails \mathsf{G}$.

\item
if $\mathsf{G}={\tt f}$, then the assumption $P\entails \mathsf{G}$ is false and the implication holds.

\item
if $\mathsf{G}=\mathsf{G}_1 \wedge \mathsf{G}_2$ we take as inductive hypotheses that
$$
\begin{array}{l}
\forall R,S.\ R \sim_{\mathsf{F}} S \wedge R\entails \mathsf{G}_1 \Rightarrow S \entails \mathsf{G}_1\\
\forall R,S.\ R \sim_{\mathsf{F}} S \wedge R\entails \mathsf{G}_2 \Rightarrow S \entails \mathsf{G}_2
\end{array}
$$
We need to prove that $Q\entails \mathsf{G}$.
Since $P\entails \mathsf{G}=\mathsf{G}_1 \wedge \mathsf{G}_2$ we have $P\entails \mathsf{G}_1$ and $P\entails \mathsf{G}_2$.
Since $P\sim_{\mathsf{F}} Q$, by inductive hypotheses we get $Q\entails \mathsf{G}_1$ and $Q\entails \mathsf{G}_2$.
Hence $Q\entails \mathsf{G}_1 \wedge \mathsf{G}_2=\mathsf{G}$.

\item
if $\mathsf{G}=\mathsf{G}_1 \vee \mathsf{G}_2$  we take as inductive hypotheses that
$$
\begin{array}{l}
\forall R,S.\ R \sim_{\mathsf{F}} S \wedge R\entails \mathsf{G}_1 \Rightarrow S \entails \mathsf{G}_1\\
\forall R,S.\ R \sim_{\mathsf{F}} S \wedge R\entails \mathsf{G}_2 \Rightarrow S \entails \mathsf{G}_2
\end{array}
$$
We need to prove that $Q\entails \mathsf{G}$.
Since $P\entails \mathsf{G}=\mathsf{G}_1 \vee \mathsf{G}_2$ we have $P\entails \mathsf{G}_1$ or $P\entails \mathsf{G}_2$.
If $P\entails \mathsf{G}_1$, since $P\sim_{\mathsf{F}} Q$, by inductive hypotheses we get $Q\entails \mathsf{G}_1$ and thus $Q\entails \mathsf{G}_1 \vee \mathsf{G}_2=\mathsf{G}$.
If $P\entails \mathsf{G}_2$, since $P\sim_{\mathsf{F}} Q$, by inductive hypotheses we get $Q\entails \mathsf{G}_2$ and thus $Q\entails \mathsf{G}_1 \vee \mathsf{G}_2=\mathsf{G}$.

\item
if $\mathsf{G}=\langle\chi\rangle \mathsf{H}$ we take as inductive hypothesis that
$$
\forall R,S.\ R \sim_{\mathsf{F}} S \wedge R\entails \mathsf{H} \Rightarrow S \entails \mathsf{H}
$$
We need to prove that $Q\entails \mathsf{G}$.
Since $P\entails \langle\chi\rangle \mathsf{H}$ it means that there exists $\upsilon, P'$ such that $P\xRightarrow{\upsilon} P'$ with $\upsilon\entails \chi$ and $P'\entails \mathsf{H}$.
Since $P\sim_{\mathsf{F}} Q$, there exists $w,Q'$ such that $Q\xRightarrow{w} Q'$ with $w\entails \chi$ and $P'\sim_{\mathsf{F}} Q'$. Then, by inductive hypothesis, $Q'\entails \mathsf{H}$ and thus $Q\entails \langle\chi\rangle \mathsf{H}=\mathsf{G}$.

\item
if $\mathsf{G}=[\chi] \mathsf{H}$ we take as inductive hypothesis that
$$
\forall R,S.\ R \sim_{\mathsf{F}} S \wedge R\entails \mathsf{H} \Rightarrow S \entails \mathsf{H}
$$
We need to prove that $Q\entails \mathsf{G}$.
If there is no $\upsilon\entails \chi$ such that $Q\xRightarrow{\upsilon}Q'$ for some $Q'$, then 
 $Q\entails [\chi] \mathsf{H}=\mathsf{G}$ trivially.
For any $\upsilon,Q'$ such that $Q\xRightarrow{\upsilon}Q'$ with $\upsilon\entails \chi$, then as $P\sim_{\mathsf{F}} Q$ there must exist $w,P'$ such that $P\xRightarrow{w}P'$ with $w\entails \chi$ and $P'\sim_{\mathsf{F}} Q'$. Since $P\entails \mathsf{G}=[\chi] \mathsf{H}$ then it must be $P'\entails \mathsf{H}$. Since $P'\sim_{\mathsf{F}} Q'$, by inductive hypothesis $Q'\entails \mathsf{H}$. Hence $Q\entails [\chi] \mathsf{H} = \mathsf{G}$.
\end{itemize}

\item[$\equiv_{\mathcal{L}_{\mathsf{F}}}\ \subseteq\ \sim_{\mathsf{F}}$:]
We prove that $\equiv_{\mathcal{L}_{\mathsf{F}}}$ is a bio-simulation and thus included in $\sim_{\mathsf{F}}$.
Take two generic processes $P\equiv_{\mathcal{L}_{\mathsf{F}}} Q$ and suppose $P\xRightarrow{\upsilon}P'$ for some $\upsilon,P'$. 
\begin{itemize}
\item
If $\upsilon \entails \mathsf{F}$ we want to prove that there exists some $w,Q'$ such that $Q\xRightarrow{w}Q'$, with $w\entails \mathsf{F}$ and $P'\equiv_{\mathcal{L}_{\mathsf{F}}} Q'$.

Towards a contradiction, assume that we cannot find such $w,Q'$.

If there is no transition $Q\xRightarrow{w}Q'$ such that $w\entails \mathsf{F}$, then the bioHML formula $\mathsf{G} \defeq \langle \mathsf{F}\rangle{\tt t}$ is such that $P\entails \mathsf{G}$ and $Q\not\entails \mathsf{G}$, contradicting the assumption $P\equiv_{\mathcal{L}_{\mathsf{F}}} Q$.

Otherwise, let $\mathcal{Q}\defeq \{Q'~\mid~\exists w.\ Q\xRightarrow{w}Q' \ \wedge\ w\entails \mathsf{F}\}$ be the (non-empty) set of processes reachable from $Q$ via a transition with a (complete) label that satisfies $\mathsf{F}$. Since our processes are with guarded recursion, the set $\mathcal{Q}$ is finite. Let $\mathcal{Q} = \{Q'_1,...,Q'_n\}$. By hypothesis all processes in $\mathcal{Q}$ must not be bio-logically equivalent to $P'$, hence for any $i\in[1,n]$ there exists a bioHML formula $\mathsf{G}_i$ such that $P'\entails \mathsf{G}_i$ and $Q'_i\not\entails \mathsf{G}_i$ (if it was the opposite, $P'\not\entails \mathsf{H}_i$ and $Q'_i\entails \mathsf{H}_i$ for some $\mathsf{H}_i$, we can use the converse formula $\mathsf{G}_i\defeq \overline{\mathsf{H}}_i$). But then the formula $\mathsf{G}\defeq \langle \mathsf{F}\rangle(\mathsf{G}_1\wedge\dots\wedge \mathsf{G}_n)$ is such that $P\entails \mathsf{G}$ and $Q\not\entails \mathsf{G}$, contradicting the assumption $P\equiv_{\mathcal{L}_{\mathsf{F}}} Q$.

\item
If $\upsilon \not\entails \mathsf{F}$ then the proof is analogous to the previous case (by exploiting $\neg \mathsf{F}$) and thus omitted.
\end{itemize}
\end{description}
\end{proof}
